\documentclass[5pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{siunitx}
\usepackage{pxfonts}
\usepackage{tabto}
\usepackage{listings}
\usepackage[margin=1.5in]{geometry}
\addtolength{\topmargin}{-.5in}

\lstdefinelanguage{rippl}{
	keywords={ over, let, in, fun, if, then, else, and, or, not, true, false, pre, head, tail, int, bool, float, char}
}
\lstset{
         numbers=left,
         firstnumber=1,
  		 numberfirstline=true,
 		 xleftmargin=5.0ex,
         basicstyle=\ttfamily,
         tabsize=4,
         commentstyle=\color{mygrey},
         keywordstyle=\bfseries
 }

\title{Rippl: Recursively Inferred Pure functional Pythonic Language}
\author{Da Hua Chen, Hollis Lehv, Amanda Liu, Hans Montero}
\date{}
\begin{document}

\maketitle

\section{Introduction}

\section{Motivation}

\pagebreak

\section{Language Features}
\subsection{List Comprehensions}
\subsection{Currying and Partial Application}
\subsection{Lazy Evaluation}
\subsection{Immutability}
\subsection{Type Inference}
\pagebreak
\section{Syntax}
\begin{multicols}{2}
\noindent $e \ ::= $ \\
\hspace*{10mm} $| \quad c$ \\
\hspace*{10mm} $| \quad x$ \\
\hspace*{10mm} $| \quad e \ e$ \\
\hspace*{10mm} $| \quad \texttt{fun } x . e$ \\
\hspace*{10mm} $| \quad \texttt{let }x = e \texttt{ in } e$ \\
\hspace*{10mm} $| \quad \texttt{if } e \texttt{ then } e \texttt{ else } e$ \\
\hspace*{10mm} $| \quad \gamma$ \\

\noindent $\gamma \ ::= $ \\
\hspace*{10mm} $| \quad \texttt{[}e...\texttt{]}$ \\
\hspace*{10mm} $| \quad \texttt{[}e...e\texttt{]}$ \\
\hspace*{10mm} $| \quad \texttt{[}e \ | \ x \texttt{ over } \gamma, \texttt{]}$ \\
\hspace*{10mm} $| \quad \texttt{[}e \ | \ x \texttt{ over } \gamma, e \texttt{]}$ \\

\noindent $\sigma \ ::= $ \\
\hspace*{10mm} $| \quad \texttt{int}$ \\
\hspace*{10mm} $| \quad \texttt{float}$ \\
\hspace*{10mm} $| \quad \texttt{bool}$ \\
\hspace*{10mm} $| \quad \texttt{char}$ \\

\noindent $\tau \ ::= $ \\
\hspace*{10mm} $| \quad \texttt{[]} $ \\
\hspace*{10mm} $| \quad \texttt{()} $ \\
\columnbreak \\
$Expressions$ \\
literals \\
variables \\
application \\
lambda abstraction \\
let binding \\
if-then-else \\
list comprehension \\

\noindent $List \ Comprehensions$ \\
infinite list \\
ranged list \\
parametric list \\
qualified list \\

\noindent $Proper \ Types$ \\
integer \\
floating point number \\
boolean \\
character \\

\noindent $Higher \ Order \ Types$ \\
list \\
tuple \\
\end{multicols}

\noindent \texttt{// Single line comments will be marked by two forward slashes} \\

\noindent \texttt{/* Multiline comments will be delimited by slash-stars} \\
\texttt{ \hspace*{3mm} Similar to a C-like program */} \\

\newpage 

\section{Operators}
\subsection{Numeric}
Our numeric or $num$ type represents the typeclass that encompasses $int$ and $float$. This use of this typeclass is mainly meant to aid in inference and protect the user from explicit typecasting.
\begin{multicols}{3}
\noindent $Operator$ \\
\hspace*{5mm} + \\
\hspace*{5mm} - \\
\hspace*{5mm} * \\
\hspace*{5mm} / \\
\hspace*{5mm} \% \\
\hspace*{5mm} \^ \\
\hspace*{5mm} $>$ \\
\hspace*{5mm} $>=$ \\
\hspace*{5mm} $<$ \\
\hspace*{5mm} $<=$ \\
\hspace*{5mm} == \\
\hspace*{5mm} != \\
\columnbreak \\
\noindent $Type$ \\
$num \ \rightarrow \ num \ \rightarrow \ num $ \\
$num \ \rightarrow \ num \ \rightarrow \ num $ \\
$num \ \rightarrow \ num \ \rightarrow \ num $ \\
$num \ \rightarrow \ num \ \rightarrow \ num $ \\
$int \quad \rightarrow \ int \quad \rightarrow \ int $ \\
$num \ \rightarrow \ num \ \rightarrow \ num $ \\
$num \ \rightarrow \ num \ \rightarrow \ bool $ \\
$num \ \rightarrow \ num \ \rightarrow \ bool $ \\
$num \ \rightarrow \ num \ \rightarrow \ bool $ \\
$num \ \rightarrow \ num \ \rightarrow \ bool $ \\
$num \ \rightarrow \ num \ \rightarrow \ bool $ \\
$num \ \rightarrow \ num \ \rightarrow \ bool $ \\
\columnbreak \\
$Function$ \\
addition \\
subtraction \\
multiplication \\
division \\
modulus \\
power \\
greater than \\
greater than or equal \\
less than \\
less than or equal \\
equal \\
not equal \\
\end{multicols}
\subsection{Boolean}
Boolean values are of type $bool$ and consist of the literals \texttt{true} and \texttt{false}. 
\begin{multicols}{3}
\noindent $Operator$ \\
\hspace*{5mm} \texttt{and} \\
\hspace*{5mm} \texttt{or} \\
\hspace*{5mm} \texttt{not} \\
\columnbreak \\
\noindent $Type$ \\
$bool \ \rightarrow \ bool \ \rightarrow \ bool $ \\
$bool \ \rightarrow \ bool \ \rightarrow \ bool $ \\
$bool \ \rightarrow \ bool \ \rightarrow \ bool $ \\
\columnbreak \\
\noindent $Function$ \\
and \\
or \\
negation \\
\end{multicols}
\subsection{Character}
Character literals are delimited by single quotes and are of type \textit{char}. Strings are constructed and treated as lists of \textit{char}.
\begin{multicols}{3}
\noindent $Operator$ \\
\hspace*{5mm} == \\
\columnbreak \\
\noindent $Type$ \\
$char \ \rightarrow \ char \ \rightarrow \ bool $ \\
\columnbreak \\
\noindent $Function$ \\
equal \\
\end{multicols}
\subsection{Lists}
A list is a first order type that is parametrically polymorphic in the type of its elements. All elements of a list must be the same type. Lists as well as list types are delimited with brackets (ie. \texttt{[int]}, \texttt{[0,3,1]}).
\begin{multicols}{3}
\noindent $Operator$ \\
\hspace*{5mm} \texttt{pre} \\
\hspace*{5mm} \texttt{head} \\
\hspace*{5mm} \texttt{tail} \\
\hspace*{5mm} \texttt{len} \\
\hspace*{5mm} ++ \\
\columnbreak \\
\noindent $Type$ \\
$a \ \rightarrow \ [a] \ \rightarrow \ [a] $ \\
$[a] \ \rightarrow \ a $ \\
$[a] \ \rightarrow \ [a] $ \\
$[a] \ \rightarrow \ int $ \\
$[a] \ \rightarrow \ [a] \ \rightarrow \ [a] $ \\
\columnbreak \\
\noindent $Function$ \\
prepend \\
head \\
tail \\
length \\
concatenate \\
\end{multicols}
\subsection{Tuples}
A tuple is a first order type, parametrically polymorphic in the types of its elements. A tuple can contain two elements of any type. Tuples as well as their types will be delimited by parentheses (ie. \texttt{("PLT",4118)}, \texttt{([char],int)}).
\begin{multicols}{3}
\noindent $Operator$ \\
\hspace*{5mm} \texttt{first} \\
\hspace*{5mm} \texttt{sec} \\
\columnbreak \\
\noindent $Type$ \\
$(a,b) \ \rightarrow \  a$ \\
$(a,b) \ \rightarrow \  b$ \\
\columnbreak \\
\noindent $Function$ \\
first element of tuple \\
second element of tuple \\
\end{multicols}

\section{Sample Programs}
\subsection{Mersenne Numbers}
Mersenne numbers are numbers that are one less than a power of two. The search for large prime Mersenne numbers, called Mersenne Primes, is an ongoing distributed computing project. This small sample program uses a list comprehension to create the first $n$ Mersenne numbers.
\begin{lstlisting}[language=rippl]
mersenne :: int -> [int]
mersenne n = [2 ^ x - 1| x over [1...n]]
\end{lstlisting}
\subsection{Collatz Conjecture}
The Collatz conjecture is a conjecture in mathematics surrounding the iterative function shown below:
$$ \begin{cases} 
      \frac{1}{2}x & x \texttt{ is even} \\
      3x + 1 & x \texttt{ is odd} 
   \end{cases}
$$
Lothar Collatz proposed that this sequence will always converge to $1$ starting from an arbitrary positive integer. Given a starting integer, the following program returns a list of iterations that ends in 1. The program hence uses $1$ as its base case even though it hasn't been mathematically proven that all numbers will eventually reach this case, but we believe in Lothar.
\begin{lstlisting}[language=rippl]
collatz :: int -> [int]
collatz n =
    let rec_collatz n list =
        if n == 1
            then 1 pre list
        else if n % 2 == 0
            then rec_collatz (n / 2) (n pre list)
            else rec_collatz (3*n +1) (n pre list)
    in rec_collatz n []
\end{lstlisting}
\subsection{Weak Prime Number Theorem}
Bertrand's Postulate stated in the Weak Prime Number Theorem that there is always a prime number to be found between some $n$ and its double $2n$. This postulate was later proven by Pafnuty Chebyshev and refined by Paul Erd\"{o}s. The following program includes a function that determines the primality of a number and a function that takes a $n$ and returns the first prime between $n$ and $2n$.
\begin{lstlisting}[language=rippl]
is_prime :: int -> bool
is_prime n =
    let max = n / 2 in
    let range = [2...max] in
    let divisors = [x | x over range, n % x == 0] in
    len divisors == 0
    
prime_number_theorem :: int -> int
prime_number_theorem n =
    let range = [(n+1)...2*n] in
    let odd_range = [x | x over range, x % 2 != 0 ] in
    foldl (fun prev -> fun curr -> if is_prime then prev else curr)
        (head odd_range) odd_range
\end{lstlisting}
\end{document}
